A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-04 rigid_bodies.tex
Update of section Molecular Dynamics

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
\mathchardef\mhyphen="2D
\usepackage{booktabs}
\usepackage{multirow}

@@

\author{Ana J. Silveira}
\email{asilveira@plapiqui.edu.ar}
\affiliation{Chemical Engineering Department, Escola de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}
\affiliation{Planta Piloto de Ingenier\'ia Qu\'imica, PLAPIQUI, Universidad Nacional del Sur, Camino La Carrindanga Km 7-CC: 717, Bah\'ia Blanca, Argentina}

\author{Charlles R. A. Abreu}
\email{abreu@eq.ufrj.br}
\affiliation{Chemical Engineering Department, Escola de Quimica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}

@@

@@

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

@@

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ . By differentiating both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}

%\subsection{Euler Parameters and Unit Quaternions}
\subsection{The Rate of Change of Unit Quaternions}
\label{sec:eulerparameters}

@@

The relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.

Finally, after carrying out the needed differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
\end{align}
\end{subequations}

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\mt B_k$ is the corresponding permutation matrix among those presented in Appendix \ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

In order to rewrite Eq.~\ref{eq:EDO_system}, the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ are opened using Eq.~\ref{eq:product_B_vector}, leading to the final form for the equations of motion:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \vt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

\subsection{Symplectic Numerical Integration}  
\label{symplectic}
An exact solution for the system of equations represented by Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, it would describe a conservative dynamics with respect to the Hamiltonian $\mathcal{H}$, in which phase space volume element would be preserved. Nevertheless, a general analytical solution does not exist. This is so, above all, because $U(\vt r, \vt q)$ is arbitrary, but even in the simplest case of free motion (i.e., when $U$ is constant), no closed-form solution is known due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}. Therefore, it becomes necessary to introduce the Liouville operators formalism, which allows us to derive symplectic numeric integrators rigorously.

For a given trajectory in phase-space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i L f,
\]
where $i L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{i L t}f_0$, where $f_0$ is the initial condition for $f$. As $f$ can be a phase-space coordinate, premultiplication with the exponential operator $e^{i L t}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} normally represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on to derive a complete numerical solution. In what follows, subscript $0$ always denotes ``initial values''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0
\end{align*}
Clearly, this corresponds to a uniform linear translation. Since the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected. Of course, the solution above is exactly the effect of the propagator $e^{i L_t t}$, where the Liouville operator $i L_t$ is, simply,
\[
i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
\]

In the second hypothetical case, referred to as the acceleration case, we consider a Hamiltonian $\mathcal{H}_a = U(\vt r, \vt q)$. The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 + t \vt F_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C_0 \vt \tau_0
\end{align*}
Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well (hence the subscript $0$ assigned to them). This solution clearly conserves $\mathcal{H}_a$, and marks the effect of a propagator $e^{i L_a t}$, obtained from the operator
\[
i L_a = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
\]

Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation. The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = (1/2) \omega_k {\mt B}_k \vt q$, and $\dot{\vt \pi} = (1/2) \omega_k {\mt B}_k \vt \pi$. A special feature of these equations makes them analytically solvable. It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well. As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form. The following analytical solution takes place:\cite{Miller2002}
\begin{align}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt \pi_0
\end{align}
These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{i L_k t}$ derived from the operator
\[
i L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{{\mt B}_k} \diff{}{\vt \pi} \right).
\]

Now note that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
\begin{align}
\label{eq:full_operator}
i L_{NVE} = i L_t + i L_a + \sum_{k=1}^3 i L_k,
\end{align}

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic geometry of Eq.~\ref{eq:EDO_system_alternative}. Such formula is
\[
e^{i L_A t + i L_B t} \approx \left( e^{i L_B \frac{\Delta t}{2}} e^{i L_A \Delta t} e^{i L_B \frac{\Delta t}{2}} \right)^N + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/N$ is the time step size and $N$ is the number of steps. In the next section we show the factorization schemes we considered for generating the dynamics in both the NVE and NVT ensembles.

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^t}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

\raggedbottom

\usepackage{booktabs}
\usepackage{multirow}

@@

\author{Ana J. Silveira}
\email{asilveira@plapiqui.edu.ar}
\affiliation{Planta Piloto de Ingenier\'ia Qu\'imica, PLAPIQUI, Universidad Nacional del Sur, Camino La Carrindanga Km 7-CC: 717, Bah\'ia Blanca, Argentina}

\author{Charlles R. A. Abreu}
\email{abreu@eq.ufrj.br}
\affiliation{Chemical Engineering Department, Escola de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}

@@

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\end{abstract}

@@

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

@@

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ . By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined in Eq.~\ref{eq:operator_S}.

\subsection{The Rate of Change of Unit Quaternions}
\label{sec:eulerparameters}

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Note that the product $\mt C \vt \tau$ results in a quaternion orthogonal to $\vt q$.

Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B} \vt \pi = \frac{1}{2} {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
\end{align}
\end{subequations}

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

In order to rewrite Eq.~\ref{eq:EDO_system}, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to a new form for the equations of motion:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \vt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

\subsection{Symplectic Numerical Integration}  
\label{symplectic}

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. However, no closed-form solution is known, even in the simplest case of free motion (i.e., when $U$ is constant), due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}. Nevertheless, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010}

For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on to derive a complete approximate solution. In what follows, subscript $0$ denotes ``initial values''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0
\end{align*}

This corresponds to a free translation (hence the subscript $t$). As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected. Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
\[
i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
\]

In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$. The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 + t \vt F_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C_0 \vt \tau_0
\end{align*}

Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well, which justifies the subscript $0$ assigned to them. This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
\[
i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
\]

Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation. The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$. A special feature of this system facilitates its solution. It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well. As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form. The following analytical solution then takes place:\cite{Miller2002}
\begin{align*}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{align*}

These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
\[
i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
\]

Now note that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
\begin{align}
\label{eq:full_operator}
i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
\end{align}

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}. For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
\[
e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^N + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/N$ is the time step size and $N$ is the number of steps. We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side. Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations. Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.
